Follow-up Samples

### 2017-03-12 submission/manuscript_outline_20161024.tex

\usepackage{pdfpages}
\usepackage{comment}

@@

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. This study tested the hypothesis that changes in the
bacterial community occur after treatment for adenoma or carcinoma.
Specifically, we tried to identify components within the community that
were different before and after removal of lesion (adenoma or
carcinoma).

@@

In this study we tested the hypothesis that there are detectable changes
to the bacterial community between pre- (initial) and post- (follow up)
treatment of lesion. We analyzed changes in alpha and beta diversity as
well as the relative abundance of specific Operational Taxonomic Units
(OTUs). We utilized Random Forest to build two models: The first was
built to classify lesion versus non-lesion (normal) while the second was
built to classify initial versus follow up samples. Subsequent
observations on how these models and OTUs within them performed before
and after treatment helped inform us as to whether initial and follow up
samples were changing and whether it was towards a normal community. We
also investigated the two models for similar important OTUs to identify
the crucial OTUs involved with both classifying lesion or normal and
initial versus follow up samples. This study will provide evidence as to
whether treatment can influence the community and if the CRC microbiome,
identified in previous studies, persists.

@@

\textbf{\emph{Bacterial Community and FIT:}} Within our 67 person cohort
we first wanted to test whether there were any broad differences between
initial and follow up samples based on lesion being either adenoma (n =
41) or carcinoma (n = 26). We found that the bacterial community in
those with carcinoma were more dissimilar to their initial sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. We also
found that there were larger changes in fecal blood (measured by FIT)
for those with carcinoma versus adenoma (P-value \textless{} 0.0001)
{[}Figure 1B{]}. The bacterial community structure before and after
surgery was visualized using NMDS for both adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05) and carcinoma {[}Figure 1D{]} (PERMANOVA
\textless{} 0.05). Interestingly, when initial and follow up samples
were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). When investigating alpha diversity
metrics there was no difference found between initial and follow up
samples for lesion, adenoma only, or carcinoma only for any metric
tested {[}Table S1{]}. We also observed that there was no difference in
the relative abundance of any OTU between initial and follow up samples
for lesion, adenoma only, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}9{]}. So we next examined whether there were changes in
previously well described carcinoma associated OTUs, such as
\emph{Porphyromonas asaccharolytica} (Otu000202), \emph{Fusobacterium
nucleatum} (Otu000442), \emph{Parvimonas micra} (Otu001273), and
\emph{Peptostreptococcus stomatis} (Otu001682). We first observed that
only a small percentage of those with adenoma or carcinoma were positive
or had a relative abundance above 0.1\% for any of these respective OTUs
{[}Figure 2{]}. Despite this, those with carcinoma had a decrease in
relative abundance from initial to follow up for \emph{Parvimonas micra}
(P-value \textless{} 0.05) and \emph{Porphyromonas asaccharolytica}
(P-value \textless{} 0.05) {[}Figure 2A{]}. In contrast, there was no
difference in relative abundance in any of these OTUs when considering
only those with adenoma {[}Figure 2B{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
and carcinoma versus normal controls. We investigated this by creating a
model to classify samples as lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.692 -
0.761 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3A{]}. There were
a total of 54 variables that were used in this model {[}Figure 3B{]}.
The FIT measurement for fecal blood resulted in the largest decrease in
MDA while the OTU with the largest MDA was \emph{Lachnospiraceae}
(Otu000015) {[}Figure 3B{]}.

If there were common OTUs that could separate adenoma and carcinoma from
normal controls, we would expect to find a decrease in the positive
probability of the follow up sample to be a lesion. This is what we
observed for the lesion model (P-value \textgreater{} 0.001). When we
separated individuals based on whether they had an adenoma or carcinoma
there was only a decrease in positive probability for the carcinoma
group {[}Figure 3C{]} (P-value \textgreater{} 0.001) and not for the
adenoma group {[}Figure 3D{]} (P-value \textgreater{} 0.05). We also
observed that there were no significant differences between the
predicted and actual calls (P-value \textless{} 0.05). The lesion model
was also able to correctly classify the one individual who still had a
carcinoma on follow up {[}Figure 3C{]}.

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we built a separate model specifically to be
able to classify whether samples were initial (before lesion was
removed) samples based on the bacterial community and FIT measurements.
The initial sample model had an AUC range of 0.657 to 0.796 after 100
iterations of 20 repeated 10-fold cross validations. The test set AUC
range for this model performed better then the training set AUCs. There
was a marked decrease in the ROC curve for the final model used when
compared to the 100 test set AUC iterations {[}Figure 4A{]}. There were
a total of 70 variables that were used for this model {[}Figure 4B{]}.
The variable that resulted in the largest MDA was \emph{Ruminococcaceae}
(Otu000278) while FIT measurement for fecal blood resulted in the sixth
largest decrease in MDA {[}Figure 4B{]}.

If there were common OTUs that could separate initial from follow up
sample regardless of whether the lesion was adenoma or carcinoma we
would expect to find a decrease in the positive probability of the
follow up sample to be an initial sample. This is what we observed for
the initial sample model (P-value \textless{} 0.001). When we separated
individuals based on whether they had an adenoma or carcinoma there was
a decrease in positive probability for both the carcinoma group
{[}Figure 4C{]} (P-value \textless{} 0.001) and for the adenoma group
{[}Figure 4D{]} (P-value \textless{} 0.001). For this model there was no
difference between the predicted and actual classifications (P-value
\textgreater{} 0.05).

\textbf{\emph{Common OTUs to both Models:}} We next wanted to compare
the similarity between the OTU variables used in either model. The main
purpose was to identify which OTUs were important not only for the
classification of lesion but also for the classification of initial or
follow up sample. Potentially, these specific OTUs are the most
important with respect to the bacterial microbiome response to removal
of lesion. When we compared the two different models with each other
there were a total of 32 common OTUs. Some of the most common taxonomic
identifications belonged to \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}. The
vast majority of these OTUs had classifications to bacteria typically
thought of as commensal {[}Table S2{]}.

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). This suggests that follow up samples for those
treated for chemotherapy my have had a larger change from the initial
sample then those without such treatment. All other variables that were
tested showed no difference based on whether chemotherapy or radiation
was used {[}Table S3{]}.

@@

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}7,8,10--12{]} by exploring what happens to the bacterial community
after lesion removal. Based on previous work by Arthur, et al. {[}13{]}
it may not be surprising to have E.coli as one of the most important
OTUs and one that was common to both models. Interestingly, many of the
most important OTUs had taxonomic identification for resident gut
microbes. This could suggest that the bacterial community is one of the
first components that could change during the pathogenesis of disease.
These bacterial microbiome changes could be the first step in allowing
more inflammatory bacterium to gain a foothold within the colon {[}9{]}.

From our results there were large observed differences in the bacterial
microbiome in samples before and after lesion removal based on whether
the individual had an adenoma or carcinoma. In individuals with
carcinoma compared to adenoma there were much larger differences between
initial and follow up samples based on the community beta diversity and
in fecal blood as measured by FIT {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for any alpha
diversity metric measured regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure S1{]}.

Although there were no differences when investigating all OTUs, when
looking specifically at four OTUs that taxonomically classified to
previously suggested carcinoma associated microbes we found that only
2/4 had a decrease in relative abundance between initial and follow up
for those with carcinoma and 0/4 had differences for those with adenoma.
This data would suggest that these specific OTUs may be important in the
transition of an adenoma to a carcinoma but less so in the initiation of
an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma and carcinoma). Based
on this model we found that the follow up samples were closer to normal
then the initial samples due to a decrease in positive probability for
lesion and that the commonly associated CRC bacteria were not highly
represented within this model with the exception of \emph{Porphyromonas
asaccharolytica} {[}Figure 3{]}. Although there was a detectable change
towards what would be expected for normal controls, it should be noted
that the follow up samples may not be a completely normal bacterial
microbiome.

After creating the lesion mode we then created a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples suggesting
that regardless of adenoma or carcinoma there are distinct common
changes within the bacterial microbiome that occurs after lesion removal
{[}Figure 4{]}. Both models had OTUs that overwhelmingly belonged to
commensal bacteria. Providing additional information on the importance
of commensal bacteria was that there were a total of 32 OTUs in common
to both models and the vast majority belonged to regular residents of
our gut community {[}Table S2{]}.

Within our study there was a significant difference for the time elapsed
in the collection of the follow up sample between adenoma and carcinoma
(uncorrected P-value \textless{} 0.05), with time passed being less for
adenoma (253 +/- 41.3 days) than carcinoma (351 +/- 102 days). These
results would indicate that the findings described were specific to the
surgical intervention and that some of the differences observed between
carcinoma and adenoma samples could be due to differences in collection
time between samples for the two different groups. Specifically, it
could confound the observation that carcinomas changed more than
adenomas {[}Figure 1A \& 1D{]}. This confounding though would not affect
the observations where these individuals were grouped together {[}Figure
3 \& 4{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after removal {[}Figure 3C \&
3D{]}. It should be noted that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that they are not the main bacteria changing after
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after lesion removal {[}Figure
2{]}.

Many of the common OTUs between the two models had OTUs that
taxonomically classified to potential butyrate producers {[}Table S2{]}.
Another batch of OTUs classified to bacteria that can either degrade
polyphenols or are inhibited by them. Both butyrate and polyphenols are
thought to be protective against cancer in part by reducing inflammation
{[}14{]}. These protective compounds are derived from the breakdown of
fiber, fruits, and vegetables by resident gut microbes. One example of
this potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}15{]}. High protein consumption in general has been linked with an
increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}17{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}18{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Interestingly, within the initial sample model the test data performed
better than the training data. This may have occurred because the
training AUC determined from 20 repeated 10 fold cross validation
removed samples at random and did not take into account that they were
matched samples. Another potential reason is that the model itself may
be over-fit since the total number of samples was not that large.
However, the lesion model did not suffer from these discrepancies.
Further independent studies need to be carried out to verify our
findings since we are dealing with correlations that may not be truly
representative of the pathogenesis of disease.

Despite these limitations our findings add to the existing scientific
knowledge on CRC and the bacterial microbiome: That there is a
measurable difference in the bacterial community after adenoma and
carcinoma removal. Further, the ability for machine learning algorithms
to take bacterial microbiome data and successfully lower positive
probability after adenoma and carcinoma removal provides evidence that
there are specific signatures, mostly attributable to commensal
organisms, associated with these lesions. Our data provides evidence
that commensal bacteria may be important in the development of polyps
and also potentially the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}7{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize back to a pre-colonscopy
community within this time period {[}19{]}. Our follow up data set had a
total of 67 individuals that not only had a sample as described but also
a follow up sample between 188 - 546 days after lesion removal and
treatment had been completed. This study was approved by the University
of Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}20{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}21{]}. DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with normal, adenoma, and carcinoma evenly
represented on each run. Each group was randomly assigned to avoid
biases based on sample collection location.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}21{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}22{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our follow up data set of 67 individuals. It should be
noted that all individuals with an adenoma or carcinoma were grouped
together to form the lesion group and the model was not created to find
differences between normal, adenoma, and carcinoma but rather
differences between both adenoma and carcinoma versus normal.

In brief, the model included data on FIT and the bacterial microbiome.
Non-binary data was checked for near zero variance and OTUs that had
near zero variance were removed. This pre-processing was performed with
the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
involved taking the samples and making 100 different 80/20 (train/test)
splits of the data where normal and lesion were represented in the same
proportion within both the whole data set and the 80/20 split. Each of
these splits were then run through 20 repeated 10-fold cross validations
to optimize the mtry hyper-parameter by maximizing the AUC (Area Under
the Curve of the Receiver Operator Characteristic). This resulting model
was then tested on the 20\% of the data that was originally held out
from this overall process. Next, in order to assess which variables were
most important to the model we counted the number of times a variable
was present in the top 10\% of mean decrease in accuracy (MDA) for each
of the 100 different 80/20 split models and then filtered this list to
variables that were only present more than 50\% of the time. This final
collated list of variables was what was considered the most important
for the model. This reduced data set was then run through the mtry
optimization again. Once the ideal mtry was found the entire 423 sample
set was used to create the final Random Forest model on which
classifications on the 67-person cohort was completed.

@@

\textbf{\emph{Initial Sample Model Creation:}} We also investigated
whether a model could be created that could identify pre (initial) and
post (follow up) lesion removal samples from each other. The main
difference was that only the 67-person cohort was used at all stages of
model building and classification. Other than this difference the
creation of this model and optimization of the mtry hyper-parameter was
completed using the same procedure as was used for the lesion model.
Instead of classifying samples as positive or negative of lesion this
model classified samples as positive or negative for being an initial
sample prior to lesion removal.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}23{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

>>>>

\usepackage{pdfpages}
\usepackage{comment}
\usepackage{lscape}

@@

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. This study tested the hypothesis that treatment for
adenoma or carcinoma result in changes to the bacterial community.
Specifically, we tried to identify components within the community that
were different before and after removal of lesion (adenoma or
carcinoma).

@@

In this study we tested the hypothesis that there are detectable changes
to the bacterial community between pre- (initial) and post- (follow up)
treatment of lesion. We analyzed changes in alpha and beta diversity as
well as the relative abundance of specific Operational Taxonomic Units
(OTUs). We then investigated several important genera from oral microbes
that have been suggested to be important in CRC pathogenesis {[}9{]}. We
next utilized Random Forest to build two models: The first was built to
classify lesion versus non-lesion (normal) while the second was built to
classify initial versus follow up samples. Subsequent observations on
how these models and OTUs within them performed before and after
treatment helped inform us as to whether initial and follow up samples
were changing and whether it was towards a normal community. We also
investigated the two models for similar important OTUs to identify the
crucial OTUs involved with both classifying lesion or normal and initial
versus follow up samples. Finally, since treatment also included
chemotherapy and radiation therapy along with lesion removal, we tested
if these additionaly treatments significantly lowered any of the metrics
examined versus those who received only lesion removal. This study will
provide evidence as to whether treatment can influence the community and
if the CRC microbiome, identified in previous studies, persists.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma (n = 41) or carcinoma (n = 26) had any
broad differences between their initial and follow up samples. We found
that those with carcinoma had a more dissimilar bacterial community
between their initial and follow up sample than those with adenoma
(P-value \textless{} 0.001) {[}Figure 1A{]}. The bacterial community
structure before and after surgery was visualized using NMDS for both
adenoma {[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05) and carcinoma
{[}Figure 1C{]} (PERMANOVA \textless{} 0.05). Interestingly, when
initial and follow up samples were compared, regardless of whether the
lesions were adenoma or carcinoma, there was no significant overall
difference in beta diversity (PERMANOVA \textgreater{} 0.05). There was
no difference between initial and follow up samples when investigating
alpha diversity metrics for lesion, adenoma only, or carcinoma only for
any metric tested {[}Table S1{]}. Additionally, there was also no
difference in the relative abundance of any OTU between initial and
follow up samples for lesion, adenoma only, or carcinoma only {[}Figure
S1{]}.

\textbf{\emph{Carcinoma Associated Genera:}} For carcinoma {[}Figure
S2A{]} (P-value \textgreater{} 0.05) but not adenoma {[}Figure S2B{]}
(P-value \textgreater{} 0.05) there was a significant decrease in
\emph{Porphyromonas}, \emph{Parvimonas}, and \emph{Peptostreptococcus}
between initial and follow up samples {[}Figure S2{]}.
\emph{Fusobacterium} for both adenoma and carcinoma had no significant
differences between initial and follow up samples {[}FIgure S2{]}
(P-value \textgreater{} 0.05). Although there were significant
differences, only a small percentage of those with adenoma or carcinoma
were positive for any of these specific genera {[}Figure S2{]}.

\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
were more normal than initial samples by building a model that
classified lesion versus normal. This model had an AUC range of 0.692 -
0.761 after 100 iterations of 20 repeated 10-fold corss validations. The
ROC curve for the final lesion model used was within the observed range
of the 100-different test set AUC iterations {[}Figure 2A{]}. There was
a total of 54 OTUs that were used in this model {[}Figure 2B{]}.
\emph{Lachnospiraceae} (Otu000015) resulted in the largest decrease in
MDA {[}FIgure 2B{]}. Heavily represented genera within the model
included OTUs from \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Roseburia}, \emph{Blautia}, and \emph{Ruminococcus}.

If there was movement towards a more normal community after treatment of
lesion, we would expect to find a decrease in the positive probability
of the follow up sample to be a lesion. We observed no such decrease in
positive probability for either those with adenoma or carcinoma
{[}FIgure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
significant difference between the predicted and actual calls (P-value
\textless{} 0.05). This model correctly classified the one individual
who still had carcinoma on follow up and this individual's positive
probability of lesion increased between their initial and follow up
sample {[}Figure 2C{]}.

\textbf{\emph{The Treatment Model:}} To test if there were differences
that could be used to classify initial and follow up treatment samples
we built a model to classify initial samples. This treatment model had
an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better than the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
were used for this model {[}Figure 3B{]}. The variable that resulted in
the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
Similarly, to the lesion model heavily represented genera included OTUs
from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, and \emph{Ruminococcus}.

If there were changes between intial and follow up samples we would
expect the positive probability of being an initial sample to be
decreased in the follow up. The is what we observed for the treatment
model (P-value \textless{} 0.001). When we separated lesion into adenoma
and carcinoma there was a decrease in positive arobability for both the
carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
{[}FIgure 3D{]} \textless{} 0.001). For this model, there was no
difference between the predicted and actual classifications (P-value
\textgreater{} 0.05).

\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
predictors within the lesion model were also in the initial sample
model. The main purpose was to identify which TOUs could be important in
lesion formation and were impacted by treatment. When we compared the
two different models with each other there were a total of 32 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
were common between models had classifications to bacteria typically
thought of as commensal {[}Table S2{]}.

\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
these changes due to treatment we assessed whether chemotherapy or
radiation impacted the observed results. Only the treatment model had a
significant change in positive probability for those treated with
chemotherapy or radiation versus those who received neither {[}Figure 4A
\& Table S3{]} (P-value \textless{} 0.05). Using the 32 common OTUs we
tested how chemotherapy or radiation affects important members of the
community. FOr both chemotherapy or radiation, the directionaly of
change of these OTUs were similar {[}Figure 4B \& 4C{]}. After multiple
comparison correction, only \emph{Blautia} (Otu000006) remained
significantly different between those who did and did not receive
chemotherapy. This data suggests that follow up samples from those
treated with either chemotherapy or radiation may have had a larger
change from initial sample due to affects on the bacterial community.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,10--12{]}. Here we describe how the bacterial community changes
after treatment and which OTUs may be most important to lesion
classification. Interestingly, many of the most important OTUs for both
treatment and lesion classification had taxonomic identification for
resident gut microbes which could suggest that members within the
commensal community may be the first that change during pathogenesis of
disease. Further, these changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}9{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
identifying commonalities within both adenoma and carcinoma groups
before and after treatment. Although there were differences for genera
associated with specific bacterium linked with CRC {[}Figure S2{]}.
These changes were not consistent across lesion. Instead, most of our
results provide support that the first members of the community to
change and potentially stay changed even after treatment are those that
are commensal bacteria {[}Figure 2-3 \& Table S3{]}. Additionally,
treatment with chemotherapy or radiation changed these bacteria and
provided a larger decrease in positive probability then removal of
lesion.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are several reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from other OTUs. Third, it is also
possible that changes in commensal bacteria are the common differences
across adenoma and carcinoma. Finally, it is also possible that all
these explanations could have played a role. Our observations though,
would suggest that an individual's resident bacteria have a large role
to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of wither lesion or
initial sample before treatment {[}Figure 2 \& 3{]}. It should be noted
that our study does not argue against the importance of these CRC
associated bacteria in the pathogenesis of disease but rather that they
are not the main bacteria changing after treatment. It is possible that
these CRC associated bacteria are important in the transition from
adenoma to carcinoma and would be one explanation as to why in our data
we not only see high initial relative abundances in carcinoma and not
adenoma individuals but also large decreases in relative abundance in
some of those with carcinoma but not in those with adenoma after lesion
removal {[}Figure S2{]}.

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers {[}Table S2{]}. Another batch of OTUs
classified to bacteria that can either degrade polyphenols or are
inhibited by them. Both butyrate and polyphenols are thought to be
protective against cancer in part by reducing inflammation {[}13{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}14{]}. High protein consumption in general has been linked with an
increased CRC risk {[}15{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}16{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma and carcinoma (uncorrected P-value \textless{} 0.05), with time
passed being less for adenoma (253 +/- 41.3 days) than carcinoma (351
+/- 102 days). These results would indicate that some of the differences
observed between the carcinoma and adenoma groups could be due to
differences in collection time. Specifically, it could confound the
observation that carcinomas changed more than adenomas {[}Figure 1A \&
1C{]}. This confounding though would not affect the observations where
these individuals were grouped together {[}Figures 2-4{]}.

Another limitation was that we do not know whether individuals who were
still classified as positive by the lesion model eventually had a
subsequent CRC diagnosis. This information would help to strengthen the
case for this model keeping several individuals above the cutoff
threshold even though at follow up they were diagnosed as no longer
having a lesion. This study also drew heavily from those with Caucasian
ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur.

Interestingly, within the treatment model the test data performed better
than the training data. This may have occurred because the training AUC
determined from 20 repeated 10-fold cross validation removed samples at
random and did not consider that they were matched samples. Another
potential reason is that the model itself may have been over-fit since
the total number of samples was not that large. However, the lesion
model did not suffer from these discrepancies. Further independent
studies need to be carried out to verify our findings on lesion and
treatment changes due to these limitations.

Despite these shortcomings our findings add to the existing scientific
knowledge on CRC and the microbiome: That there is a measurable
difference in the bacterial community after adenoma and carcinoma
treatment. Further, the ability for machine learning algorithms to take
OTU data and successfully lower positive probability of lesion or
initial sample after treatment provides evidence that there are specific
signatures, mostly attributable to commensal organisms, associated with
both treatment and lesion. Our data provides evidence that commensal
bacteria may be important in the development of polyps and potentially
the transition from adenoma to carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}7{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonscopy community within this time period
{[}17{]}. Our study cohort consisted of 67 individuals with an initial
sample as described and a follow up sample obtained between 188 - 546
days after treatment of lesion. This study was approved by the
University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}18{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The initial and follow up samples were sequenced on
the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}18{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}19{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main training and testing of the model
completed on an independent data set of 423 individuals. This model was
then applied to our 67-person cohort. It should be noted that all
individuals with an adenoma or carcinoma were grouped together to form
the lesion group and the model was not created to find differences
between normal, adenoma, and carcinoma but rather differences between
both adenoma and carcinoma versus normal.

The model included only OTU data obtained from 16S rRNA sequencing.
Non-binary data was checked for near zero variance and OTUs that had
near zero variance were removed. This pre-processing was performed with
the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
involved making 100 different 80/20 (train/test) splits of the data
where normal and lesion were represented in the same proportion within
both the whole data set and the 80/20 split. For each a 20 repeated
10-fold cross validation was performed on 80\% component to optimize the
mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). The resulting model was then tested
on the hold out data obtained from the 20\% component. Assessment of the
most important OTUs to the model involved counting the number of times
an OTU was present in the top 10\% of mean decrease in accuracy (MDA)
for each of the 100 different splits run. This was then followed with
filtering of this list to variables that were only present in more than
50\% of these 100 runs. The final collated list of variables was then
run through the mtry optimization again. Once the ideal mtry was found
the entire 423 sample set was used to create the final Random Forest
model on which classifications on the 67-person cohort was completed.

@@

\textbf{\emph{Treatment Model Creation:}} We also investigated whether a
model could be created that could identify pre- (initial) and post-
(follow up) treatment samples. The main difference was that only the
67-person cohort was used at all stages of model building and
classification. Other than this difference the creation of this model
and optimization of the mtry hyper-parameter was completed using the
same procedure as was used for the lesion model. Instead of classifying
samples as positive or negative of lesion this model classified samples
as positive or negative for being an initial sample prior to treatment.

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}20{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}21{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.
